PyPy's Approach to Virtual Machine Construction

### 2006-06-01 talk/dls2006/paper.tex
Minor tweaks on the paper.

The front-end part of the translation process analyses the input
RPython program in two phases, as follows:\footnote{Note that the two
phases are intermingled in time, because type inference proceeds from
an entry point function and follows all calls, and thus only gradually
discovers (the reachable parts of) the input program.}
%
\begin{enumerate}
\item We take as input RPython function objects,\footnote{The input to our
       translation chain are indeed loaded runtime function objects,
       not source code or ASTs.  This enables a form of staged
       programming in which we can use unrestricted
       Python for meta-programming purposes at load time: the whole of
       the source program -- as Python program -- produces the RPython
       program input to the tool-chain as the object graph loaded in
       memory.  This includes both the relevant functions and prebuilt
       data.} and convert them to control flow graphs -- a structure
       amenable to analysis.  These flow graphs contain polymorphic
       operations only: in Python, almost all operations are
       dynamically overloaded by type.

@@

We have two alternatives, each implemented as a transformation step.
The first one inserts naive reference counting throughout the whole
program's graphs, which without further optimizations gives exceedingly
bad performance (it should be noted that the CPython interpreter is also
based on reference counting, and experience suggests that it was not a
bad choice in this particular case).

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
and replaces all \texttt{malloc} operations with calls to a garbage
collector's allocation routine.  It can inspect all the graphs to discover the
\texttt{struct} types in use by the program, and assign a unique type id to
each of them.  These type ids are collected in internal tables that
describe the layout of the structures, e.g.\ their sizes and the location
of the pointer fields.

We have implemented other transformations as well, e.g.\ performing
various optimizations, or turning the whole code into a style that
allows us to use coroutines (still in ANSI C: it is a variant of
continuation-passing that will be the subject of another paper.)
Another example is the exception transformer, which transforms graphs
that still contain implicit exception handling into a form suitable for
C (currently based on a global flag to signal the presence of an
exception, set and checked around selected function calls).

Finally, currently under development is a variant of the very first
transformation step, for use when targeting higher-level,
object-oriented (OO) environments.  It is currently being designed
together with back-ends for Smalltalk/Squeak\footnote{Our simple OO
type system is designed for \textit{statically-typed} OO environments,
including Java; the presence of Smalltalk as a back-end might be
misleading in that respect.} and CLI/.NET.  This first transformation
step, for C-like environments, is called the \textit{LLTyper} or
low-level typer: it produces
C-level flow graphs, where the object-oriented features of RPython
(classes and instances) become manipulations of C structs with
explicit virtual table pointers.  By contrast, for OO environments the
transformation step is called the \textit{OOTyper}: it targets a simple
object-oriented type system, and preserves the classes and instances
of the original RPython program.  The LLTyper and OOTyper still have
much code in common, to convert the more Python-specific features like
its complex calling conventions.

@@

This means that in addition to transforming the existing graphs, each
transformation step also needs to insert new functions into the forest.
A key feature of our approach is that we can write such ``system-level''
code -- relevant only to a particular transformation -- in plain Python
as well.  The idea is to feed these new Python functions into the
front-end, using this time the transformation's target (lower-level)
type system during the type inference.  In other words, we can write
plain Python code that manipulates objects that conform to the
lower-level type system, and have these functions automatically
transformed into appropriately typed graphs.

@@

In the example of the \texttt{malloc} operation, replaced by a call to
GC code, this GC code can invoke a complete collection of dead
objects, and can thus be arbitrarily complicated.  Still, our GC code
is entirely written in plain Python, and it manipulates ``objects'' that
are still at a lower level: pointer and address objects.  Even with
the restriction of having to use pointer-like and address-like
objects, Python remains more expressive than, say, C to write a GC
(the work on the Jikes RVM's GC \cite{JikesGC} was the inspiration to try to
express GCs in Python, see section \ref{relatedwork}).

@@

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared to other approaches
(e.g.\ Squeak \cite{Squeak}), we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system.\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}

@@

The other type systems, however, do not correspond to built-in Python
types.  For each of them, we implemented:
%
\begin{enumerate}
\item The types, which we use to tag the variables of the graphs at
      the given level (types are actually annotated, self-recursive
      formal terms, and would have been implemented simply as such if
      Python supported them directly).

\item The Python objects that emulate instances of these types (more
      about them below).
\end{enumerate}
%
We have defined well-typed operations between instances of these types,
syntactically expressed with standard Python operators (e.g.\ if \texttt{x}
is a C-level array, \texttt{x[n]} accesses its \texttt{n}th item).
The emulating instances provide a concrete implementation of
these operations that works in normal Python; the types involved in the
operations are also known to the type inference engine when it
analyses system code like the helpers of figure \ref{fig_llappend}.

@@

The third reason is fundamental: we use these emulating objects to
\textit{represent} pre-built objects at that level.  For example, the GC
transformer instantiates the objects emulating C arrays for the internal
type id tables, and it fills them with the correct values.  These array
objects are then either used directly when testing the GC, or translated
by the C back-end into static pre-initialized arrays.

@@

Remember that building the control flow graphs is not done, as one might
first expect, by following a function at the syntactic level.  Instead,
the whole program is imported in a normal Python interpreter; the full
Python language is used at this point as a kind of preprocessor with
meta-programming capabilities.  Once the program is imported, the object
data in memory consists of Python function objects in bytecode format,
and any other kind of objects created at import-time, like class
objects, prebuilt instances of those, prebuilt tables, and so on.  Note
that these objects have typically no text representation any more; for
example, cyclic data structures may have been built at this point.  The
translation tool-chain first turns these function objects into in-memory
control flow graphs which contain direct references to the prebuilt data
objects, and then handles and transforms these graphs.

@@

The actual transformation from function objects -- i.e.\ bytecode -- to
flow graph is performed by the Flow Object Space, a short but generic
plug-in component for the Python interpreter of PyPy.  The architecture
of our Python interpreter is shown in figure \ref{interpobjspace}.

\begin{figure}
  \centering
  \label{interpobjspace}
  \begin{tabular}{|c|c|} \hline
\multicolumn{2}{|c|}{\vphantom{\Large X}forest of bytecode objects from the application}    \\ \hline
\multicolumn{2}{|c|}{\vphantom{\Large X}Python bytecode interpreter}            \\ \hline
    \vphantom{\Large X} Standard Object Space     &  Flow Object Space  \\ \hline
  \end{tabular}
  \caption{the interpreter and object spaces.}
\end{figure}

Note that the left column, i.e.\ the bytecode interpreter and the
Standard Object Space, form the full Python interpreter of PyPy.  It is
an RPython program, and the whole purpose of the translation process is
to accept this as \textit{input}, and translate it to an efficient form.
The description of this particular input program is beyond the scope of
the present paper; see \cite{S}.

@@

\begin{figure*}
\centering
\includegraphics[scale=0.75]{image/flowlattice.pdf}
\caption{the lattice order of the flow object space.}
\label{flowlattice}
\end{figure*}

@@

Note that we produce flow graphs in Static Single Information or SSI\cite{SSI}
form, an extension of Static Single Assignment\cite{SSA}: each variable is
only used in exactly one basic block.  All variables that are not dead
at the end of a basic block are explicitly carried over to the next
block and renamed.

@@

Naturally, simply propagating types forward requires the use of a fixed
point algorithm in the presence of loops in the flow graphs or in the
inter-procedural call graph.  Indeed, we flow types forward from the
beginning of the entry point function into each basic block, operation
after operation, and follow all calls recursively.  During this process,
each variable along the way gets a type.  In various cases, e.g.\ when we
close a loop, the previously assigned types can be found to be too
restrictive.  In this case, we generalise them to allow for a larger set
of possible run-time values, and schedule the block where they appear
for reflowing.  The more general type can generalise the types of the
results of the variables in the block, which in turn can generalise the
types that flow into the following blocks, and so on.  This process
continues until a fixed point is reached.

@@

\item $NullableStr$, $NullableInst(class)$ -- a string or
      \texttt{None}; resp. an instance or \texttt{None}.
\end{itemize}
%
Figures \ref{latticeoverall} and \ref{latticedetail} show how these
types are ordered to form a lattice.  We mostly use its structure of
join-semilattice only.

@@

All list terms for all variables are unordered.  The Pbcs form a
classical finite set-of-subsets lattice.  In addition, we have left
out a number of other annotations that are irrelevant for the basic
description of the annotator and straightforward to handle:
$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
complete list is described in \cite{T}.

The type system moreover comes with a family of rules, which for every
operation and every sensible combination of input types describes the
type of its result variable.  Let $V$ be the set of Variables that
appear in the user program's flow graphs.  Let $b$ be a map from $V$
to $A$; it is a ``binding'' that gives to each variable a type.  The
purpose of the annotator is to compute such a binding stepwise.

@@

It is outside the scope of the present paper to describe the type
inference engine and the rules more formally.  The difficult points
involve mutable containers -- e.g.\ initially empty list that are
filled somewhere else -- and the discovery of instance attributes --
in Python, classes do not declare upfront a fixed set of attributes
for their instances, let alone their types.  Both these examples
require sophisticated reflowing techniques that invalidate existing
types in already-annotated basic blocks, to account for the influence
of more general types coming indirectly from a possibly distant part
of the program.  The reader is referred to technical report \cite{D}
for more information.

@@

We give formal termination and correctness proofs in \cite{D}, as well as
worst-case bounds and some experimental evidence of their practical
irrelevance.

\subsection{Precision}

@@

As described in the introduction, the most important downside of our
approach is that automatic specialization is a potential
performance-killer.  So we decided that in the normal case, all calls to
a given function contribute to a single annotation of that function.
Inside the function graph, we propagate the \textit{join} of the types
of the actual parameters of the call sites.  We \textit{do} support
specialization, however: we can generate several independently-annotated
copies of the flow graphs of certain functions.  When annotating RPython
programs, such specialization does not happen automatically: we rely on
hints provided by the programmer in the source code, in the form of
flags attached to function objects.  As we had this trade-off in mind
when we wrote the Python interpreter of PyPy, we only had to add a dozen
or so hints in the end.

@@

At the level of RPython, on the other hand, the range of specializations
that make sense is much wider.  We have used anything between
specialization by the type of an argument, to specialization by an
expected-to-be-constant argument value, to memoized functions that the
type inference engine will actually call during annotation and replace
by look-up tables, to complete overriding of the annotator's behavior in
extreme cases.  In this sense, the need for manual specialization turned
into an advantage, in term of simplicity and flexibility of implementing
and using new specialization schemes.

This conclusion can be generalized.  We experimented with a simple
approach to type inference that works well in practice, and that can
very flexibly accomodate changes in the type system and even completely
different type systems.  We think that the reasons for this success are
to be found on the one hand in the (reasonable) restrictions we put on
ourselves when designing the RPython language and writing the Python
interpreter of PyPy in RPython, and on the other hand in an ad-hoc type
system that is designed to produce enough precision (but not more) for
the purpose of the subsequent transformations to C-level code.

>>>>

The front-end part of the translation process analyses the input
RPython program in two phases, as follows:\footnote{Note that the two
phases are intermingled in time, because type inference proceeds from
an entry point function and follows all calls, and thus only gradually
discovers (the reachable parts of) the input program.}
%
\begin{enumerate}
\item We take as input RPython function objects,\footnote{The input to our
       translation chain are indeed loaded runtime function objects,
       not source code nor ASTs.  This enables a form of staged
       programming in which we can use unrestricted
       Python for meta-programming purposes at load time: the whole of
       the source program -- as a Python program -- produces the RPython
       program which is sent to the tool-chain in the form of its object
       graph loaded in
       memory.  This includes both the relevant functions and prebuilt
       data.} and convert them to control flow graphs -- a structure
       amenable to analysis.  These flow graphs contain polymorphic
       operations only: in Python, almost all operations are
       dynamically overloaded by type.

@@

We have two alternatives, each implemented as a transformation step.
The first one inserts naive reference counting throughout the whole
program's graphs, which without further optimizations gives a somewhat
bad performance (it should be noted that the CPython interpreter is also
based on reference counting, and experience suggests that it was not a
bad choice in this particular case).

The other, and better, alternative is an exact GC, coupled with a
transformation, the \textit{GC transformer}.  It inputs C-level-typed graphs
and replaces all \texttt{malloc} operations with calls to a garbage
collector's allocation routine.
The transformation inspects all the graphs to discover the
\texttt{struct} types in use by the program, and assigns a unique type id to
each of them.  These type ids are then collected in internal tables that
describe the layout of the structures, e.g.\ their sizes and the location
of the pointer fields.

We have implemented other transformations as well, e.g.\ performing
various optimizations, or turning the whole code into a style that
allows us to use coroutines (still in ANSI C: it is a variant of
continuation-passing that will be the subject of another paper.)
Another example is the exception transformer, which transforms graphs
that still contain implicit exception handling into a form suitable for
C (currently based on a global flag to signal the presence of an
exception, which is set and checked around selected function calls).

Finally, currently under development is a variant of the very first
transformation step, for use when targeting higher-level,
object-oriented (OO) environments.  It is currently being designed
together with back-ends for Smalltalk/Squeak\footnote{Our simple OO
type system is designed for \textit{statically-typed} OO environments,
including Java; the presence of Smalltalk as a back-end might be
misleading in that respect.} and CLI/.NET.  This first transformation
step, for C-like environments, is called the \textit{LLTyper} or
low-level typer: it produces
C-level flow graphs, where the object-oriented features of RPython
(classes and instances) become manipulations of C structs with
explicit virtual table pointers.  By contrast, for OO environments the
transformation step is called the \textit{OOTyper:} it targets a simple
object-oriented type system, and preserves the classes and instances
of the original RPython program.  The LLTyper and OOTyper still have
much code in common, to convert the more Python-specific features like
its complex calling conventions.

@@

This means that in addition to transforming the existing graphs, each
transformation step also needs to insert new functions into the forest.
A key feature of our approach is that we can write such ``system-level''
code -- relevant only to a particular transformation -- in plain Python
as well.  The idea is to feed these new Python functions into the
front-end, using this time the transformation's target (lower-level)
type system during the type inference.  In other words, we can write
plain Python code that manipulates objects that conform to the
lower-level type system, and turn these functions into graph that are
directly typed in this lower-level type system.

@@

In the example of the \texttt{malloc} operation, replaced by a call to
GC code, this GC code can invoke a complete collection of dead
objects, and can thus be arbitrarily complicated.  Still, our GC code
is entirely written in plain Python, and it manipulates ``objects'' that
are still at a lower level: pointer and address objects.  Even with
the restriction of having to use pointer-like and address-like
objects, Python remains more expressive than, say, C to write a GC
(the work on the Jikes RVM's GC \cite{JikesGC} was the inspiration to try to
express GCs in Python; see section \ref{relatedwork}).

@@

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared with other approaches
(e.g.\ Squeak \cite{Squeak}), we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system.\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}

@@

The other type systems, however, do not correspond to built-in Python
types.  For each of them, we implemented:
%
\begin{enumerate}
\item The types, which we use to tag the variables of the graphs at
      the given level (types are actually annotated self-recursive
      formal terms, and would have been implemented simply as such if
      Python supported them directly).

\item The Python objects that emulate instances of these types (more
      about them below).
\end{enumerate}
%
We have defined well-typed operations between instances of these types,
syntactically expressed with standard Python operators (e.g.\ if \texttt{x}
is a C-level array, \texttt{x[n]} accesses its \texttt{n}th item).
The emulating instances provide a concrete implementation of
these operations that works in normal Python; the types involved in the
operations are also known to the type inference engine when it
analyses system code like the helper of figure \ref{fig_llappend}.

@@

The third reason is fundamental: we use these emulating objects to
\textit{represent} pre-built objects at that level.  For example, the GC
transformer instantiates an object emulating a C array for the internal
type id table, and fills it with the correct values.  This array
object is then either used directly when testing the GC, or translated
by the C back-end into a static pre-initialized array.

@@

Remember that building the control flow graphs is not done, as one might
first expect, by following a function at the syntactic level.  Instead,
the whole program is imported in a normal Python interpreter -- the full
Python language is used as a kind of preprocessor with
meta-programming capabilities.  Once the program is imported, the object
data in memory consists of Python function objects in bytecode format,
as well as any other kind of objects created at import-time, like class
objects, prebuilt instances of those, prebuilt tables, and so on.  Note
that these objects have typically no text representation any more; for
example, cyclic data structures may have been built at this point.  The
translation tool-chain first turns these function objects into in-memory
control flow graphs which contain direct references to the prebuilt data
objects, and then handles and transforms these graphs.

@@

\begin{figure}
  \centering
  \begin{tabular}{|c|c|} \hline
\multicolumn{2}{|c|}{\vphantom{\Large X}forest of bytecode objects from the application}    \\ \hline
\multicolumn{2}{|c|}{\vphantom{\Large X}Python bytecode interpreter}            \\ \hline
    \vphantom{\Large X} Standard Object Space     &  Flow Object Space  \\ \hline
  \end{tabular}
  \caption{the interpreter and object spaces.}
  \label{fig_interpobjspace}
\end{figure}

The actual transformation from function objects -- i.e.\ bytecode -- to
flow graph is performed by the Flow Object Space, a short but generic
plug-in component for the Python interpreter of PyPy.  The architecture
of our Python interpreter is shown in figure \ref{fig_interpobjspace}.
Note that the left column, i.e.\ the bytecode interpreter and the
Standard Object Space, form the full Python interpreter of PyPy.  It is
an RPython program, and the whole purpose of the translation process is
to accept this as \textit{input}, and translate it to an efficient form.
The description of this particular input program is beyond the scope of
the present paper; see \cite{S}.

@@

\begin{figure*}
\centering
\includegraphics[scale=0.667]{image/flowlattice.pdf}
\caption{the lattice order of the flow object space.}
\label{flowlattice}
\end{figure*}

@@

Note that we produce flow graphs in Static Single Information or SSI\cite{SSI}
form, an extension of Static Single Assignment\cite{SSA}: each variable is
only used in exactly one basic block.  All variables that are not dead
at the end of a basic block are explicitly carried over to the next
block and renamed, as can be seen in figure \ref{fig_interpobjspace}:
each link carries a list of variables matching the formal input
arguments of its target block.

@@

Naturally, simply propagating types forward requires the use of a fixed
point algorithm in the presence of loops in the flow graphs or in the
inter-procedural call graph.  Indeed, we flow types forward from the
beginning of the entry point function into each basic block, operation
after operation, and follow all calls recursively.  During this process,
each variable along the way gets a type.  In various cases, e.g.\ when we
close a loop, the previously assigned types can be found to be too
restrictive.  In this case, we generalise them to allow for a larger set
of possible run-time values, and schedule the block where they appear
for reflowing.
The newly generalised types can in turn generalise the types of other
result variables in the block, which in turn can generalise the
types that flow into the following blocks, and so on.  This process
continues until a fixed point is reached.

@@

\item $NullableStr$, $NullableInst(class)$ -- a string or
      \texttt{None}; resp. an instance or \texttt{None}.
\end{itemize}

Figures \ref{latticeoverall} and \ref{latticedetail} show how these
types are ordered to form a lattice -- we generally use its structure of
join-semilattice only, but we will see in section \ref{precision}
a use case for the \textit{meet}.
Not shown on the figures are the list terms, which are simply
unordered with each other, and the Pbcs, which form a
classical finite set-of-subsets lattice.  In addition, we have left
out a number of other annotations that are irrelevant for the basic
description of the annotator and straightforward to handle:
$Dictionary$, $Tuple$, $Float$, $UnicodePoint$, $Iterator$, etc.  The
complete list is described in \cite{T}.

@@

The type system moreover comes with a family of rules, which for every
operation and every meaningful combination of input types describes the
type of its result variable.  Let $V$ be the set of Variables that
appear in the user program's flow graphs.  Let $b$ be a map from $V$
to $A$; it is a ``binding'' that gives to each variable a type.  The
purpose of the annotator is to compute such a binding stepwise.

@@

It is outside the scope of the present paper to describe the type
inference engine and the rules more formally.  The difficult points
involve mutable containers, e.g.\ initially empty list that are
filled somewhere else; the discovery of instance attributes --
in Python, classes do not declare upfront a fixed set of attributes
for their instances, let alone their types; and the discovery of
base class interfaces -- across a class hierarchy, some methods with
the same name may be called polymorphically by the program, while
others may be unrelated methods used internally by the subclasses.
These examples require reflowing techniques that invalidate existing
types in already-annotated basic blocks, to account for the influence
of more general types coming indirectly from a possibly distant part
of the program.  The reader is referred to \cite{D}
for more information.

@@

We give formal termination and correctness proofs in \cite{D}, as well as
worst-case bounds and common-case estimates.

\subsection{Precision}
\label{precision}

@@

As described in the introduction, the most important downside of our
approach is that automatic specialization is a potential
performance-killer.  So we decided that in the normal case, all calls to
a given function contribute to a single annotation of that function.
Inside the function graph, we propagate the \textit{join} of the types
of the actual parameters of the call sites.  We \textit{do} support
specialization, however: we can generate several independently-annotated
copies of the flow graph of certain functions.  When annotating RPython
programs, such specialization does not happen automatically: we rely on
hints provided by the programmer in the source code, in the form of
flags attached to function objects.  As we had this trade-off in mind
when we wrote the Python interpreter of PyPy, we only had to add about
a dozen hints in the end.

@@

At the level of RPython, the range of specializations
that make sense is actually much wider.  Although we only specialize a
very small subset of the functions, we use criteria as diverse as
specialization by the type of an argument, specialization by an
expected-to-be-constant argument value, memoized functions that the
type inference engine will actually call during annotation and replace
by look-up tables, and even complete overriding of the annotator's behavior in
extreme cases.  In this sense, the need for manual specialization turned
into an advantage, in term of simplicity and flexibility of implementing
and using new specialization schemes.

This conclusion can be generalised.  We experimented with a simple
approach to type inference that works well in practice, and that can
very flexibly accomodate small and big changes in the type system.
We think that the reasons for this success are
to be found on the one hand in the (reasonable) restrictions we put on
ourselves when designing the RPython language and writing the Python
interpreter of PyPy in RPython, and on the other hand in an ad-hoc type
system that is designed to produce enough precision (but not more) for
the purpose of the subsequent transformations to C-level code.
